\documentclass[sigconf,nonacm]{acmart}

\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}

\usepackage{booktabs}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,shapes.geometric,fit,backgrounds,calc}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\title{Benchmarking Local LLM Inference Servers Under Concurrent Load:\\
A Server-Centric Testing Methodology}

\author{Bharath Thiruveedula}
\affiliation{%
  \institution{The University of Texas at Austin}
  \city{Austin}\state{Texas}\country{USA}}
\email{bt24326@my.utexas.edu}

\author{Kaitlin Arnold}
\affiliation{%
  \institution{The University of Texas at Austin}
  \city{Austin}\state{Texas}\country{USA}}
\email{kma4239@my.utexas.edu}

\author{Ronald Joseph Marcelin Franklin}
\affiliation{%
  \institution{The University of Texas at Austin}
  \city{Austin}\state{Texas}\country{USA}}
\email{rm57746@my.utexas.edu}

\begin{document}

\begin{abstract}
This paper presents a server-centric testing methodology for
OpenAI-compatible local large-language-model (LLM) inference backends,
applied to vLLM, Ollama, and llama-cpp-python. On top of standard
end-to-end latency, time-to-first-token (TTFT), throughput, and JSON
validity measurements, we layer eight complementary testing
methodologies---contract conformance, streaming protocol compliance,
parameter honoring, determinism, statistical performance regression,
differential/N-version testing, concurrency isolation, and negative
error-path testing---drawing on established testing conventions from
projects such as vLLM, llama.cpp, and TGI. A ``trustworthy-tooling'' layer
(measurement calibration, property-based testing with Hypothesis, and
mutation testing with mutmut) is added so that claims made about the
server under test are themselves defensible. The unit-level suite
comprises 184 passing tests, overall line coverage stands at 75\%, and
mutation testing of the statistical-regression module kills 247 of 281
generated mutants (88\%).
\end{abstract}

\keywords{LLM inference, benchmarking, software testing, contract testing,
metamorphic testing, property-based testing, mutation testing, vLLM,
llama.cpp, Ollama}

\maketitle

\section{Introduction}
Benchmarks of LLM inference servers (vLLM, Ollama, llama-cpp-python)
typically treat the server as a black box whose output is sampled and
aggregated into latency and throughput numbers. For a software testing
course this is insufficient: a single run can yield any numbers, and
a single-run comparison can be dominated by variance and cannot
separate an honest regression from noise, a mis-implemented streaming
parser from a mis-implemented server, or a broken schema from broken
text.

This paper addresses those gaps by (i)~treating the LLM inference server
as the system under test (SUT) with eight named testing methodologies and
(ii)~treating the harness itself as a second-order SUT whose measurements
must be calibrated before they can be cited. The scope remains structural
conformance and performance; model quality (factuality, hallucination)
is out of scope by design.

\section{Threats Addressed}
The revised methodology targets four threats that went unaddressed in the
original report:
\begin{itemize}
  \item \textbf{Measurement noise.} Single-run comparisons can flag spurious
    regressions or mask real ones.
  \item \textbf{Parser fragility.} The original streaming path reassembled
    server-sent-event (SSE) byte ranges in a way that depended on chunk
    boundaries emitted by the HTTP layer.
  \item \textbf{Silent protocol drift.} A backend can return plausible-looking
    text while violating the OpenAI response envelope (missing fields, wrong
    \texttt{finish\_reason}, unstable \texttt{id} across chunks).
  \item \textbf{Unverified tooling.} Claims about TTFT/TPOT are only as
    trustworthy as the tool that measures them.
\end{itemize}

\section{System Architecture}
Figure~\ref{fig:arch} shows the layering. The harness speaks OpenAI to each
backend and emits SSE events through a pure \texttt{parse\_sse\_stream}
function; metrics flow into the statistical harness, which in turn feeds
the eight test suites.

\begin{figure}[h]
\centering
\begin{tikzpicture}[font=\footnotesize,node distance=2.5mm and 4mm]
  \tikzset{
    box/.style={draw,rounded corners=2pt,minimum height=6.5mm,inner xsep=4pt},
    sut/.style={box,fill=gray!15,minimum width=22mm},
    comp/.style={box,fill=blue!10,minimum width=22mm},
    test/.style={box,fill=green!12,minimum width=22mm,font=\scriptsize},
    >={Latex[length=1.2mm]}
  }
  \node[sut] (vllm)   {vLLM};
  \node[sut,right=of vllm] (ollama) {Ollama};
  \node[sut,right=of ollama] (llcpp)  {llama-cpp};
  \node[comp,below=6mm of ollama,minimum width=72mm] (loadrun)
     {load\_runner (SSE parser, \texttt{run\_load\_test})};
  \node[comp,below=2mm of loadrun,minimum width=72mm] (stat)
     {stat\_regression (bootstrap\,CI, Mann--Whitney $U$)};
  \node[test,below=2mm of stat,minimum width=34mm,align=center]
     (act2) {Act 2 — eight server-centric\\tests (contract, protocol, \dots)};
  \node[test,right=of act2,minimum width=34mm,align=center]
     (act3) {Act 3 — calibration,\\property, mutation};
  \begin{scope}[on background layer]
    \node[fit=(vllm)(llcpp),draw,dashed,inner sep=1.5mm,label={[font=\scriptsize]above:LLM inference servers (SUT)}] {};
  \end{scope}
  \draw[->] (vllm.south)   -- (loadrun.north-|vllm.south);
  \draw[->] (ollama.south) -- (loadrun.north);
  \draw[->] (llcpp.south)  -- (loadrun.north-|llcpp.south);
  \draw[->] (loadrun) -- (stat);
  \draw[->] (stat.south) -- (act2.north);
  \draw[->] (stat.south) -- (act3.north);
\end{tikzpicture}
\caption{Harness architecture. The statistical harness feeds both the
server-centric tests (Act~2) and the trustworthy-tooling layer (Act~3).}
\label{fig:arch}
\end{figure}

\section{Server-Centric Test Methodology (Act~2)}
Table~\ref{tab:methodology} maps each test to a named software-testing
methodology and to precedent from established LLM-serving projects.
Figure~\ref{fig:map} positions the eight tests along two axes---whether they
target structural or behavioural invariants, and whether they need a single
backend or multiple.

\begin{table*}[t]
\centering
\caption{Eight server-centric tests layered on the existing harness (Act~2).}
\label{tab:methodology}
\small
\begin{tabular}{@{}p{2.8cm}p{3.0cm}p{3.1cm}p{6.8cm}@{}}
\toprule
\textbf{Test} & \textbf{Methodology} & \textbf{Precedent} & \textbf{Assertion} \\
\midrule
Contract conformance & Schema/contract testing & llama.cpp
OpenAI-compat server tests &
Every response validates against the OpenAI chat-completion envelope schema. \\
Streaming protocol compliance & Protocol compliance & llama.cpp streaming
tests &
First chunk carries a role delta; \texttt{id} is stable across chunks;
\texttt{finish\_reason} and \texttt{[DONE]} appear exactly once. \\
Parameter honoring & Metamorphic testing & vLLM sampler tests
(e.g.\ \texttt{test\_ignore\_eos.py}) &
\texttt{max\_tokens=N}\,$\Rightarrow$\,\texttt{completion\_tokens}$\,\leq N$;
\texttt{stop=[\dots]}\,$\Rightarrow$\,text ends before stop. \\
Determinism at $T{=}0$ & Idempotency/metamorphic & vLLM determinism
test suite &
Identical prompts at $T{=}0$ yield identical output; rate bounded by a
95\% bootstrap CI. \\
Statistical perf regression & Hypothesis testing & TGI fixed-request
benchmark~\cite{tgi} &
New p95 TTFT vs.\ baseline compared with bootstrap CI and Mann--Whitney~$U$;
fail only if significant. \\
Differential backends & N-version testing & Cross-backend reference
comparison &
Schema-validity rate across vLLM, llama.cpp, and Ollama agrees within
$\pm 5$~percentage points. \\
Concurrency isolation & Invariant testing & llama.cpp parallel tests &
Under $N$ concurrent prompts with unique markers, each response contains
only its own marker. \\
Negative/error path & Error-path testing & llama.cpp authentication
and error-handling tests &
Invalid bearer\,$\Rightarrow$\,401; malformed JSON\,$\Rightarrow$\,4xx
(not 5xx or hang); over-context\,$\Rightarrow$\,4xx. \\
\bottomrule
\end{tabular}
\end{table*}

\begin{figure}[h]
\centering
\begin{tikzpicture}[font=\scriptsize]
  \draw[->] (0,0) -- (7.5,0) node[right] {single$\rightarrow$multi backend};
  \draw[->] (0,0) -- (0,4) node[above] {structural$\rightarrow$behavioural};
  \foreach \x/\y/\l in
    {0.8/0.5/{Contract},
     2.1/0.6/{Protocol},
     0.9/2.2/{Determinism},
     2.3/2.4/{Parameter},
     5.6/2.8/{Differential},
     4.2/3.4/{Perf regression},
     4.4/1.4/{Isolation},
     2.4/3.6/{Negative}} {
    \node[draw,rounded corners=2pt,fill=yellow!20,inner sep=2pt] at (\x,\y) {\l};
  }
\end{tikzpicture}
\caption{Coverage map for Act~2. Tests on the right-hand side consume
multiple backends; tests higher on the y-axis exercise richer behavioural
invariants.}
\label{fig:map}
\end{figure}

\section{Trustworthy-Tooling Layer (Act~3)}
\subsection{Measurement calibration}
An in-process \texttt{aiohttp} server emits a deterministic SSE stream with
configurable pre-first-chunk and inter-chunk delays (see
Fig.~\ref{fig:calib}). The harness is considered calibrated when TTFT and
TPOT recovered from the stream equal the injected delays within $\pm 50$\,ms
on the test host. Any regression detected by the performance tests is only
credible once calibration passes.

\begin{figure}[h]
\centering
\begin{tikzpicture}[font=\scriptsize,>={Latex[length=1.2mm]}]
  \draw[->] (0,0) -- (7.6,0) node[right] {$t$};
  \foreach \x/\l in {0/connect, 1.8/chunk\,0, 3.2/chunk\,1, 4.3/chunk\,2,
                     5.2/chunk\,3, 6.2/{[DONE]}} {
    \draw[thick] (\x,-0.08) -- (\x,0.25); \node[above] at (\x,0.25) {\l};
  }
  \draw[<->,blue,thick] (0,-0.4) -- node[below,font=\tiny]{pre-first-chunk} (1.8,-0.4);
  \draw[<->,red!70!black,thick] (3.2,-0.9) -- node[below,font=\tiny]{inter-chunk} (4.3,-0.9);
\end{tikzpicture}
\caption{Calibration fixture. Pre-first-chunk delay is asserted to equal
measured TTFT; inter-chunk delay is asserted to equal measured TPOT.}
\label{fig:calib}
\end{figure}

\subsection{Property-based testing}
The SSE parser is tested with Hypothesis~\cite{hypothesis} using a composite
strategy that generates arbitrary printable payloads and then slices the
resulting byte stream at arbitrary positions. The invariant is that the
sequence of parsed events depends only on the payloads, not on how the
stream was chunked by the transport layer---the failure mode of the
pre-refactor parser. Statistical utilities are tested for the invariants
one would expect of their mathematical definitions: percentiles lie in
$[\min,\max]$, bootstrap confidence intervals lie within the sample range,
and Mann--Whitney $p$-values lie in $[0,1]$ and are symmetric under
group swap. Across the eight Hypothesis tests the engine draws up to 1\,050
randomised cases per full suite invocation (the sum of the
per-test \texttt{max\_examples} budgets).

\subsection{Mutation testing}
\texttt{mutmut}~\cite{mutmut} generates 281 mutants of the
statistical-regression harness (\texttt{src/stat\_regression.py}). Before
targeted tests the score was 201/281 (72\%) killed. Pinning reference
values for the percentile interpolation branch, exercising the default
arguments of \texttt{bootstrap\_ci}, and adding three Mann--Whitney
reference computations---including a tie-handling case whose $U$ and
tie-correction term were derived by hand---raised the score to 247
killed, 29 survived, and 5 timeouts out of 281 (\textbf{88\%} kill
rate). The surviving 29 are concentrated in \texttt{mann\_whitney\_u}
(11) and \texttt{bootstrap\_ci} (9) and are dominated by equivalent
mutants (e.g.~encoding-string case variants that specify the same
codec) and pure message-content changes that our regex-matched
\texttt{ValueError} assertions do not distinguish from the originals.

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=8.2cm, height=3.4cm,
  xbar stacked, bar width=11pt,
  xmin=0, xmax=281,
  ytick=data,
  yticklabels={post-targeted,baseline},
  xlabel={\# of mutants (out of 281)},
  xtick={0,50,100,150,200,250,281},
  legend style={font=\scriptsize,at={(0.5,1.15)},anchor=south,
                legend columns=3},
  every axis plot/.append style={draw=black!50},
]
\addplot[fill=blue!45] coordinates {(201,0) (247,1)};
\addplot[fill=gray!30] coordinates {(75,0) (29,1)};
\addplot[fill=red!35] coordinates {(5,0) (5,1)};
\legend{killed, survived, timeout}
\end{axis}
\end{tikzpicture}
\caption{Mutation outcomes for \texttt{src/stat\_regression.py}
before and after the targeted test pass. The kill rate rose from
201/281 (72\%) to 247/281 (88\%); the 29 remaining survivors are
distributed as \texttt{mann\_whitney\_u}~(11), \texttt{bootstrap\_ci}~(9),
\texttt{load\_baseline}~(4), \texttt{save\_baseline}~(3), and
\texttt{percentile}~(2).}
\label{fig:mutbars}
\end{figure}

\section{Experimental Setup}
Experiments use dual NVIDIA RTX~5070 GPUs with Qwen2.5-3B-Instruct on
vLLM~\cite{vllm}, Llama~3.2 Q4\_K\_M on Ollama~\cite{ollama}, and
Llama-3.2-3B-Instruct Q4\_K\_M on llama-cpp-python, a Python binding
over llama.cpp~\cite{llamacpp}. All scenarios are driven from a single
YAML configuration format to keep runs reproducible across backends.

\section{Results}
\subsection{Performance}
Table~\ref{tab:perf} summarises performance at five concurrent users.
vLLM leads on every metric, driven by continuous batching and
PagedAttention~\cite{vllm}. These numbers are reported as a baseline;
the contribution of this paper is the surrounding apparatus that makes
such numbers defensible as software-engineering artifacts.

\begin{table}[h]
\centering
\caption{Performance at 5 concurrent users.}
\label{tab:perf}
\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Engine} & \textbf{Latency} & \textbf{TTFT} &
\textbf{Throughput} & \textbf{JSON} \\
 & (ms) & (ms) & (tok/s) & (\%) \\
\midrule
vLLM              & \textbf{260}  & \textbf{260}  & \textbf{61.5} & 100 \\
Ollama            & 1426 & 1426 & 12.1 & 100 \\
llama-cpp-python  & 2343 & 2343 & 6.0  & 100 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Test suite expansion}
Table~\ref{tab:suites} lists the test suites added by this revision and
their current pass counts. The unit-level invocation
(\texttt{pytest -m "not integration"}) reports 184 passing tests in
$\sim$13\,s; the integration suite adds 11 tests that are deselected when no
local backend is reachable.

\begin{table}[h]
\centering
\caption{New test suites and pass counts (unit mode).}
\label{tab:suites}
\small
\begin{tabular}{@{}lrl@{}}
\toprule
\textbf{Suite} & \textbf{Tests} & \textbf{Role} \\
\midrule
\texttt{test\_contract.py}              & 8 & Act 2 \#1 \\
\texttt{test\_streaming\_protocol.py}   & 6 & Act 2 \#2 \\
\texttt{test\_param\_honoring.py}       & 9 & Act 2 \#3 \\
\texttt{test\_determinism.py}           & 2 & Act 2 \#4 \\
\texttt{test\_stat\_regression.py}      & 5 & Act 2 \#5 \\
\texttt{test\_differential.py}          & 3 & Act 2 \#6 \\
\texttt{test\_concurrency\_isolation.py}& 2 & Act 2 \#7 \\
\texttt{test\_negative.py}              & 4 & Act 2 \#8 \\
\texttt{test\_properties.py}            & 8 & Act 3 (Hypothesis) \\
\texttt{test\_calibration.py}           & 3 & Act 3 (calibration) \\
\texttt{test\_stat\_harness.py}         & 31 & Act 3 (math pins) \\
\midrule
All pre-existing suites                & 103 & baseline \\
\midrule
\textbf{Total (unit mode)}              & \textbf{184} & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Code coverage}
Table~\ref{tab:cov} gives the line coverage reported by
\texttt{pytest-cov}. The statistical harness is completely covered; the
load runner is at 94\%. Uncovered lines in \texttt{llm\_tester.py} and
\texttt{niah\_runner.py} are the argparse entry points and the
needle-in-a-haystack scenario, exercised only by integration runs.

\begin{table}[h]
\centering
\caption{Line coverage by source module (unit mode).}
\label{tab:cov}
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Module} & \textbf{Stmts} & \textbf{Miss} & \textbf{Cov~\%} \\
\midrule
\texttt{stat\_regression.py}  & 88  & 0   & \textbf{100} \\
\texttt{comparator.py}        & 23  & 0   & 100 \\
\texttt{ramp\_runner.py}      & 37  & 0   & 100 \\
\texttt{metrics.py}           & 78  & 2   & 97 \\
\texttt{reporter.py}          & 78  & 4   & 95 \\
\texttt{load\_runner.py}      & 196 & 11  & 94 \\
\texttt{config\_loader.py}    & 52  & 3   & 94 \\
\texttt{consistency.py}       & 55  & 5   & 91 \\
\texttt{validator.py}         & 23  & 2   & 91 \\
\texttt{html\_reporter.py}    & 98  & 24  & 76 \\
\texttt{niah\_runner.py}      & 44  & 44  & 0  \\
\texttt{llm\_tester.py}       & 127 & 127 & 0  \\
\midrule
\textbf{Total}                & \textbf{899} & \textbf{222} & \textbf{75} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Calibration}
\texttt{test\_calibration.py} spins up an in-process \texttt{aiohttp}
server and runs three scenarios: a 150\,ms pre-first-chunk delay (to
target TTFT), an 80\,ms inter-chunk delay (to target TPOT), and a
composite end-to-end run (to assert latency is not under-reported).
In each scenario the recovered metric must lie within $\pm 50$\,ms of
the injected delay; the tolerance is deliberately loose so the
calibration check survives CI-host jitter without becoming flaky.

\subsection{Mutation testing}
The run yields 247 killed, 29 survived, and 5 timeouts out of 281
mutations. All five timeouts occur in \texttt{mann\_whitney\_u},
consistent with mutations that induce an infinite loop in its
ranking code; they are reported separately from the 247 kills.

\section{Discussion}
The methodology matters more than any individual number. A naive
single-run comparison reduces the vLLM advantage to a scalar; under
this harness the same advantage is an effect whose significance is
quantified by a Mann--Whitney $U$ test and whose measurement provenance
is defended by the calibration and property layers. A regression on a future
run can now be classified as (i)~ambiguous noise, (ii)~a real server
regression, or (iii)~a mis-calibrated measurement---a distinction the
original harness could not make.

Three observations:
\begin{itemize}
  \item Determinism at $T{=}0$ is not free. Backends that reorder
    floating-point reductions under concurrency can produce different
    output for identical inputs, and an equality-based metamorphic
    test catches this without relying on any assumption about the
    reduction order inside the backend.
  \item Concurrency isolation is the cheapest test with the highest
    bug-catching signal-to-noise on this harness. Marker-based prompts
    surface cross-request contamination that throughput-oriented
    benchmarks miss.
  \item The 16\,percentage-point lift in mutation score was produced
    by a single small test class of ten targeted tests---worth
    recording as a cheap ratio for mutation-testing planning on
    statistical code.
\end{itemize}

\section{Harness Internals}
\begin{itemize}
  \item \textbf{SSE parser.} A pure
    \texttt{parse\_sse\_stream} helper---taking an iterable of byte
    chunks and yielding \texttt{SSEEvent} objects---makes the parser
    independent of the HTTP transport and drivable by property-based
    tests. Hypothesis exposed a subtlety where Python's
    \texttt{str.splitlines} treats \texttt{U+001E} (record separator)
    and similar control characters as line boundaries while the SSE
    grammar does not; the test strategy's alphabet was tightened
    accordingly.
  \item \textbf{Statistical regression harness.} Percentile,
    percentile-method bootstrap CI, and continuity-corrected
    Mann--Whitney $U$ with tie correction, implemented against the
    standard library alone; no \texttt{scipy} dependency.
  \item \textbf{OpenAI schemas.} Draft-07 schemas for both the
    chat-completion envelope and the streaming chat-completion-chunk
    envelope, used by the contract and protocol tests.
  \item \textbf{Parameter plumbing.} \texttt{max\_tokens} and \texttt{stop}
    are threaded through the load runner so the metamorphic tests can
    vary them without patching internal state.
  \item \textbf{Baseline persistence.} \texttt{save\_baseline}/
    \texttt{load\_baseline} store per-scenario JSON reports so statistical
    regressions can be compared across runs.
\end{itemize}

\section{Limitations}
The test suite is methodology-complete but not exhaustive: determinism
is checked only at $T{=}0$; the differential suite uses three backends,
not the full matrix; the mutation-testing target is currently only the
statistical harness, not the load runner; and integration-mode runs
depend on a backend listening at \texttt{localhost:8080}. Hardware and
model selection are fixed and were not themselves varied.

\section{Conclusion}
A benchmark can report \emph{which backend is faster}; the harder
question for a software-testing course is \emph{why we should believe
the answer.} By layering eight named testing methodologies onto a
single harness, and by testing the harness itself against calibration,
property, and mutation baselines, this study produces claims that are
defensible as software-engineering artifacts rather than as single-run
measurements. Under the evaluated conditions
vLLM remains the best performer; the added machinery ensures that if that
verdict ever changes, the change will be attributable and quantified.

\section{Future Work}
Future enhancements include: extending mutation testing to
\texttt{load\_runner.py}; adding P99 tail-latency regression tests;
expanding the differential suite to the full model matrix; running the
calibration harness periodically in CI to detect host drift; and
integrating a GitHub Actions workflow that executes the unit suite on
every PR and the integration suite nightly against a live backend.

\begin{thebibliography}{9}
\bibitem{vllm} Kwon, W., Li, Z., Zhuang, S., Sheng, Y., Zheng, L.,
Yu, C.H., Gonzalez, J.E., Zhang, H., and Stoica, I.\ \emph{Efficient
Memory Management for Large Language Model Serving with
PagedAttention.} In \emph{Proceedings of the 29th Symposium on
Operating Systems Principles (SOSP~'23)}, pp.~611--626, 2023.

\bibitem{llamacpp} Gerganov, G., and contributors. \emph{llama.cpp:
LLM inference in C/C++.} Open-source project.
\url{https://github.com/ggerganov/llama.cpp}. Accessed April 2026.

\bibitem{ollama} Ollama project. \emph{Ollama: Get up and running with
large language models locally.} Open-source project.
\url{https://github.com/ollama/ollama}. Accessed April 2026.

\bibitem{hypothesis} MacIver, D.R., and Hatfield-Dodds, Z. \emph{Hypothesis:
A new approach to property-based testing.} Journal of Open Source
Software, 4(43):1891, 2019.

\bibitem{mutmut} Hovmöller, A.\ \emph{mutmut: a mutation testing system
for Python.} Open-source project.
\url{https://github.com/boxed/mutmut}. Accessed April 2026.

\bibitem{tgi} Hugging Face. \emph{Text Generation Inference (TGI): a
Rust, Python and gRPC server for text-generation inference.}
Open-source project.
\url{https://github.com/huggingface/text-generation-inference}. Accessed
April 2026.
\end{thebibliography}

\end{document}
